Static Analysis by Path Focusing

### 2011-05-02 report/report.tex
report

\def\R{\mathbb{R}}
\def\Z{\mathbb{Z}}
\def\N{\mathbb{N}}
\def\Q{\mathbb{Q}}
\def\P{\mathcal{P}}
\def\ite{\textnormal{ite\ }}

@@

However, most of the time, the solution of this fixpoint equation must be
computed in a complex domain: in program analysis, this could be the state space
of the program, i.e the set of all possible states of the program. This
computation quickly becomes too costly.

@@

\begin{proposition}
If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
$$\alpha(lfp(\Phi)) \leq_A lfp(\alpha(\Phi))$$
where $lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
\end{proposition}

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, or of finite depth, meaning that there doesn't exist
any sequence $(y_i)_{i \geq 0}$ of elements of the abstract domain $A$ such as
$\forall i \geq 0,\ y_i \leq_A y_{i+1}$. Indeed, in case of a domain of infinite
depth, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

@@

@@

In $\Q^n$, the set of polyhedra is a lattice. The least element is $\perp$
(the empty polyhedron), and the greatest element is $\top$ (the whole $Q^n$).

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. There have been different
work proposing a definition of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against the loss of precision induced by widening:

@@

Throughout the report, a program will be represented by a control flow graph:
\begin{itemize}
\item $P$ is a finite set of control points.
\item $I_p$ is the (possibly empty)
set of initial values for each control point $p \in P$.
\item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
has a semantic $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the set
of possible values of $\overrightarrow{x}$. $\tau_e$ maps a set of states before
the transition to the set of states after the transition.
\end{itemize}

@@

@@

In the case of abstract domain with infinite ascending sequence, we use the
previously defined widening operator:

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such as removing the nodes of $P_w$
disconnect every cycles. For instance, one could choose $P_w$ as the set of the
heads of loops.

\subsubsection{Iteration strategies}
There exist several iteration strategies for computing the fixpoint: whatever
the order we choose for updating the different $X_p, p\in P$, the result at the
end of the analysis is correct. Yet, the precision of the result and the time
before convergence could be very different. 
There have been some work to find efficient iteration strategies
\cite{Bou92}, such as first stabilizing innermost loops, or stabilizing strongly
connected components of the graph.

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to different reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, one compute a union of convex polyhedra instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision. 
\item There is also a loss of precision each time a point in the control flow
graph has several incoming edges. This is the case for instance for
\emph{if} statements, loops, etc. In this case, the abstract value of this point
is a convex hull of several convex polyhedra, inducing an upper approximation.
Indeed, if $X_1, X_2$ are convex polyhedra,
$$X_1 \cup X_2 \subseteq X_1 \sqcup X_2$$

@@

Typically, to analyze a program with loops, the classical approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, one could directly compute an invariant for the loop:

@@

	Let $(P,E)$ be the control flow graph of the program.
	First, one choose a set of points $P_W \subseteq P$, such that the graph
	obtained after removing these points has no cycle. The widening operator
	will be applied at these points. The choice of $P_W$ can be guided by
	\cite{Bou92}.
	Then, one choose another set $P_R \subseteq P$ of nodes, satisfying the
	following properties:
	\begin{itemize}
	\item $I_p = \emptyset$ for each $p \in P \ P_R$, meaning that the initial
	points of the program are included in $P_R$.
	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
	\end{itemize}

@@

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the boolean variables such as the formula is true.
	Among these boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

\begin{enumerate}
\item Each numerical SSA variable of the control flow graph has its definition
in the $\rho$ formula.
\item For each operation in the program, we encode its semantic, expressed in
terms of the boolean and numerical SSA variables. For instance, if the CFG
contains the affectation $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
the formula. This equality expression is noted $assign(x.2)$. 
$assign(x)$ is defined for each numerical SSA-variable of the program.
\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
$c(i,j)$ express the condition that needs to be true to go through the
transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
If the transition is undeterministic, then $c(i,j) = c_{i,j}$, where
$c_{i,j}$ is a boolean variable left undeterministic.
\item For each point $p_i \notin P_R$, we set the reachability predicate 
$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
\item For each point $p_i^d \in P_R$, we set the reachability predicate 
$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
formula. We simply use $\ite$ (\emph{if-then-else}) statements, standardly defined in the
SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
depending on the incoming transition we come from. Generally, for a
$Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:

@@

	$\rho$ express the semantic of the program. We still need to create the
	formula expressing the reachability problem defined in \ref{reachability}.

	The abstract domain $D$ we use is those of convex polyhedra, so an abstract
	value $X \in D$ is simply a system of linear inequalities between the
	different SSA-variables of the program. Then, it is easy to encode the
	property $x \in X$ into an SMT-formula: one simply write the conjunction of
	these inequalities, with the name of the variables in the vector $x$.

	The formula given to the SMT-solver is:

@@

The advantage of using such an infrastructure is that the analyzer can check
programs written in different languages: C, C++, \dots, without needing to write
a specific frontend for each of them. Here, the analyzer works directly on the
internal representation (IR), which is a graph representation of
the program. The control flow graph can easily be extracted from
this representation.

>>>>

\def\R{\mathbb{R}}
\def\Z{\mathbb{Z}}
\def\N{\mathbb{N}}
\def\Q{\mathbb{Q}}
\def\P{\mathcal{P}}
\def\ite{\textnormal{ite\ }}
\def\lfp{\textnormal{\it lfp}}

@@

However, most of the time, the solution of this fixpoint equation must be
computed in a complex domain: in program analysis, this could be the state space
of the program, i.e the set of all possible states of the program. This
computation quickly becomes too costly or doesn't terminate.

@@

\begin{proposition}
If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
\end{proposition}

@@

Termination of the fixpoint computation has to be guaranteed. This termination
depends on the properties of the abstract domain:
this one should be finite, or of finite depth, meaning that there doesn't exist
any sequence $(y_i)_{i \geq 0}$ of elements of the abstract domain $A$ such as
$\forall i \geq 0,\ y_i <_A y_{i+1}$. Indeed, in case of a domain of infinite
depth, the least fixpoint
computation could run indefinitely, since the ascending sequence of elements of
$A$ is infinite, and the fixpoint in never reached.

@@

\cite{Monniaux_HOSC09} gives a more general definition of the widening operator.

@@

In $\Q^n$, the set of polyhedra ordered by inclusion is a lattice. The least element is $\perp$
(the empty polyhedron), and the greatest element is $\top$ (the whole $Q^n$).

@@

Since the domain of convex polyhedra is of infinite depth, a widening operator
is defined to ensure convergence of the technique. There have been different
work proposing a definition of the widening operator in LRA
\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
against the loss of precision induced by widening.

@@

Throughout the report, a program will be represented by a control flow graph:
\begin{itemize}
\item $P$ is a finite set of control points.
\item $I_p$ is the (possibly empty)
set of initial values for each control point $p \in P$. For an initial control
point $p$, i.e a starting point of the program, $I_p$ is not empty. It is empty
in the other case.
\item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
has a semantic $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the set
of possible values of $\overrightarrow{x}$. $\tau_e$ maps a set of states before
the transition to the set of states after the transition.
\end{itemize}

@@

\begin{figure}[h!]
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,
                    semithick]

  \node[state] (p)             [label=left:$X_{p}$]       {$p$};
  \node[state]         (p2) [above  of=p] {};
  \node[state]         (p1) [left of=p2,label=left:$X_{p'}$] {$p'$};
  \node[state]         (p3) [right of=p2] {};
  \node         (p22) [below  of=p] {};
  \node         (p12) [left of=p22] {};
  \node         (p32) [right of=p22] {};

  \path [transition] (p1) edge              node {} (p);
  \path [transition] (p2) edge              node {} (p);
  \path [transition] (p3) edge              node {} (p);
  \path [transition2] (p) edge              node {} (p12);
  \path [transition2] (p) edge              node {} (p22);
  \path [transition2] (p) edge              node {} (p32);
\end{tikzpicture}
\end{figure}

@@

In the case of an abstract domain with infinite ascending sequence, we use the
previously defined widening operator:

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such as removing the nodes of $P_w$
disconnects every cycles. For instance, one could choose $P_w$ as the set of the
heads of loops.

\subsubsection{Iteration strategies}
There exist several iteration strategies for computing the fixpoint: whatever
order we choose for updating the different $X_p, p\in P$, the result at the
end of the analysis is correct. Yet, the precision of the result and the time
before convergence could be very different. 
There have been some work to find efficient iteration strategies
\cite{Bou92}, such as first stabilizing innermost loops, or stabilizing strongly
connected components of the graph.

\subsubsection{Sources of imprecision}
 \label{multigraph}
Linear Relation Analysis is sound but incomplete: there is loss of precision
during the computation, due to different reasons:
\begin{enumerate}
\item Since an abstract domain is used to represent the set of possible states,
such as the domain of convex polyhedra, there is an upper approximation of the
set of possible states to the smallest convex polyhedra including this set. This
loss of precision is unavoidable. Still, there are techniques proposing to
compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
program point, one compute a union of convex polyhedra instead of a single
one.
\item The widening operator, used to ensure convergence of the technique,
also induces a loss of precision. The classical technique is then to compute
narrowing iterations, which often recover some precision.

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, one obtain $i \in
[0,1]$, \dots. After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better.

\item There is also a loss of precision each time a point in the control flow
graph has several incoming edges. This is the case for instance for
\emph{if} statements, loops, etc. In this case, the abstract value of this point
is a convex hull of several convex polyhedra, inducing an upper approximation.
Indeed, if $X_1, X_2$ are convex polyhedra,
$$X_1 \cup X_2 \subseteq X_1 \sqcup X_2$$

@@

Typically, to analyze a program with loops, the approach is to use the
widening operator, for example at the head of the loops, to ensure convergence.
Since this induces imprecision, another technique called \emph{acceleration}
aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
the following program, one could directly compute an invariant for the loop:

@@

	Intuitively, expanding the graph comes back to consider independently the
	different paths in the control flow graph, and thus to be more precise,
	because paths that are not feasible will be ignored. Additionally, the
	number of control points with several incoming transitions will be reduced,
	so the loss of precision due to the convex hull of polyhedra may be
	minimized.

\begin{figure}[!h]
\label{multigraph}
\centering
\begin{minipage}[c]{.19\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node[state] (n1) [below left of=n0] {};
	\node[state] (n2) [below right of=n0] {};
	\node[state] (n3) [below right of=n1] {};
	\node[state] (n4) [below left of=n3] {};
	\node[state] (n5) [below right of=n3] {};
	\node[state] (n6) [below right of=n4] {};
	\node[state] (n7) [below left of=n6] {};
	\node[state] (n8) [below right of=n6] {};
	\node[PRstate] (n9) [below right of=n7] {$q$};

  \path [transition] 
		(n0) edge              node {} (n1);
  \path [transition] 
		(n0) edge              node {} (n2);
  \path [transition] 
		(n1) edge              node {} (n3);
  \path [transition] 
		(n2) edge              node {} (n3);
  \path [transition] 
		(n3) edge              node {} (n4);
  \path [transition] 
		(n3) edge              node {} (n5);
  \path [transition] 
		(n4) edge              node {} (n6);
  \path [transition] 
		(n5) edge              node {} (n6);
  \path [transition] 
		(n6) edge              node {} (n7);
  \path [transition] 
		(n6) edge              node {} (n8);
  \path [transition] 
		(n7) edge              node {} (n9);
  \path [transition] 
		(n8) edge              node {} (n9);
\end{tikzpicture}
\end{minipage} 
$\Longrightarrow$
\begin{minipage}[c]{.19\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$p$};
	\node (n1) [below left of=n0] {};
	\node (n2) [below right of=n0] {};
	\node (n3) [below right of=n1] {};
	\node (n4) [below left of=n3] {};
	\node (n5) [below right of=n3] {};
	\node (n6) [below right of=n4] {};
	\node (n7) [below left of=n6] {};
	\node (n8) [below right of=n6] {};
	\node[PRstate] (n9) [below right of=n7] {$q$};

  \path [transition] 
		(n0) edge  [out=0, in=0]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=180, in=180]        node {} (n9);
  \path [transition] 
		(n0) edge  [out=205, in=155]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=230, in=130]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=255, in=105]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-25, in=25]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-50, in=50]            node {} (n9);
  \path [transition] 
		(n0) edge  [out=-75, in=75]            node {} (n9);
\end{tikzpicture}
\end{minipage}
\caption{One can expand the graph in the left and obtain the associate graph in
the right.}
\end{figure}
\FloatBarrier

	Let $(P,E)$ be the control flow graph of the program.
	First, one choose a set of points $P_W \subseteq P$, such that the graph
	obtained after removing these points has no cycle. The widening operator
	will be applied at these points. The choice of $P_W$ can be guided by
	\cite{Bou92}.
	Then, one choose another set $P_R \subseteq P$ of nodes, satisfying the
	following properties:
	\begin{itemize}
	\item $I_p = \emptyset$ for each $p \in P \setminus P_R$, meaning that the initial
	points of the program are included in $P_R$.
	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
	\end{itemize}

@@

	If the problem has a solution,  the SMT-solver will be in position to show
	which path in the control flow graph is selected, by giving a model: an
	assignments of the boolean variables such that the formula is true.
	Among these boolean variables, some of them are associated to a transition
	in the control flow graph:
	\begin{itemize}
	\item if the model has set to \emph{true} the boolean variable $b_e$ associated
	to the transition $e$, then the focus path goes through this transition.
	\item in the other case, the focus path doesn't go through this transition.
	\end{itemize}

@@

\begin{enumerate}
\item Each numerical SSA variable of the control flow graph has its definition
in the $\rho$ formula.
\item For each operation in the program, we encode its semantic, expressed in
terms of the boolean and numerical SSA variables. For instance, if the CFG
contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
the formula. This equality expression is noted $assign(x.2)$. 
$assign(x)$ is defined for each numerical SSA-variable of the program.
\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
$c(i,j)$ express the condition that needs to be true to go through the
transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
$c_{i,j}$ is a boolean variable left non-deterministic.
\item For each point $p_i \notin P_R$, we set the reachability predicate 
$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
\item For each point $p_i^d \in P_R$, we set the reachability predicate 
$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
formula. We simply use $\ite$ (\emph{if-then-else}) statements, standardly defined in the
SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
depending on the incoming transition we come from. Generally, for a
$\Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:

@@

	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
	formula expressing the reachability problem defined in \ref{reachability}.

	The abstract domain $D$ we use is those of convex polyhedra, so an abstract
	value $X \in D$ is simply a system of linear inequalities between the
	different SSA-variables of the program. Then, it is easy to encode the
	property $x \in X$ into an SMT-formula: one simply writes the conjunction of
	these inequalities, with the name of the variables in the vector $x$.

	To compute the reachability strating from point $p_i$, the formula given to
	the SMT-solver is:

@@

The advantage of using such an infrastructure is that the analyzer can check
programs written in various languages: C, C++, \dots, without needing to write
a specific frontend for each of them. Here, the analyzer works directly on the
internal representation (IR), which is a graph representation of
the program. The control flow graph can easily be extracted from
this representation.
